A Bayesian ranking statistic to find high-mass black holes in LIGO--Virgo data

### 2021-03-18 main.tex
Updates from Overleaf

%%%%%%---SECTIONS---%%%%%%%%%%%%
\section{Introduction}
In the 1960s, X-ray and optical observations from the Cyg-X1 binary star system led to the discovery of the first stellar-mass black hole~\cite{}. Shortly afterwards, in 1974, kinematic measurements of stars near the center of our galaxy demonstrated the presence of a supermassive black hole of mass $10^6\ \msun$ -- Sag A$^*$~\cite{}. These two detections firmly established the existence of stellar-mass
black holes $\text{M}_\text{BH} < 10^{2} \ \msun$ and
supermassive black holes, $\text{M}_\text{BH} > 10^{5} \ \msun$ in our universe. Since then, astronomers accumulated more definitive evidence for stellar-mass and supermassive black holes. However, a gap in between the mass-ranges of these two categories began to form: there is a deficiency of black holes in the intermediate-mass range $10^{2} - 10^{5}\ \text{M}_{\odot}$. These intermediate-mass black holes must have existed to form the supermassive black holes detected to exist in the early universe~\cite{}. However, we only have a handful of candidates and one confirmed intermediate-mass black hole~\cite{}. The study and discovery of intermediate-mass black holes will help bridge the gap and provide a pathway to forming supermassive black holes.

The two main proposed formation channels for Intermediate-mass black holes are the light-seed and the heavy-seed channels. The light-seed channel involves dynamic interactions of low-mass seeds, e.g. hierarchical mergers of stellar-mass black holes, runaway collisions of stars.  The heavy-seed channel involves the direct collapse of high-mass seeds, e.g. the direct collapse of giant metal-free Population III stars or early protogalaxies' collapse.  Although promising pathways, there are challenging obstacles both scenarios must overcome to pathways that can form intermediate-mass black holes.

Hierarchical mergers in active galactic nuclei migration-traps, globular and stellar clusters may be a feasible light-seed method for black holes to accumulate mass and reach the intermediate-mass range. However, when black holes merge, they can receive recoil-kicks up to $\sim$2,500 km/s. Hence, the merger remnants require their recoil kick to be less than its environment's escape velocity to undergo successive mergers. This tight constraint makes hierarchical mergers a challenging scenario. One proposed heave-seed pathway involves the direct collapse of massive, metal-free Population III stars. Typically, massive stars $> 100 \ \msun$ can eject more than half of their mass in the form of radiatively driven stellar winds. However, at zero metallicity, mass loss through stellar winds is negligible \cite{Kudritzki 2000}. Hence,  Population III stars have the opportunity to complete their life-cycles without losing a significant portion of their mass. Moreover, Population III stars with masses above $230 \ \msun$ will avoid electron-positron pair-instability. Consequently, oxygen and silicon burning will not drive an explosion, allowing the star to undergo a complete collapse to leave behind an intermediate-mass black hole remnant with a mass of $120-280 \ \msun$, dependant on the initial star mass \cite{Bond et al. 1984}. Hence, the formation of intermediate-mass black holes from Population III stars is directly proportional to the fraction of Population III stars with masses above $230 \ \msun$, a fraction that is still uncertain~\cite{Coleman Miller, Formation Scenarios for Intermediate-Mass Black Holes}.

% - Bridge between light-seed and heavy-seed formation and observation
% - Although IMBH gap isnt completly filled, observational work conducted in the past 20 years points to the existence of intermediate-mass black holes .

% - To prove the existence of IMBH we need to measure their BH's masses
% - We can measure mass by two main methods: direct kinematic, emission (reverberation mapping, luminosity, gw)
% - Discuss Candidates
% - LIGO a good way. 190521g uncertain
% - Challenges of IMBH from ground-based
% - THUS We need a better way to discriminate between glitches and high-mass signals -- BCR!

To prove the existence of intermediate-mass black holes, the scientific community needs to measure the mass of a black hole inside the IMBH range. However, measuring the mass of black holes can be challenging.

It is possible to measure the mass of nearby supermassive black holes (such as Sag A) by directly observing the dynamic motion of stars nearby the supermassive black hole. However, this dynamic measurement method is challenging to utilise for IMBHs as their sphere of influence is too small to resolve motions. Additionally, dynamic-measurements that have alluded to IMBH candidates, e.g. the candidate at the centre of the globular cluster 47 Tucanae \cite{kiziltan 2017}, can also be attributed to simpler models without a central IMBH \cite{freire 2017}.

In a similar vein, although the M-$\sigma$ and M-L relations (relations between the spheroidal bulge of a galaxy's velocity dispersion $\sigma$,  luminosity L, and its central black hole's mass M) can estimate the mass of supermassive black holes  \cite{Ferrarese & Merritt 2000 and Gebhardt et al. 2000}, it cannot accurately estimate the mass of IMBHs. The M-$\sigma$ and M-L relations are not calibrated for low-mass galaxies that might host an IMBH as its central black hole \cite{J. E. Greene and L. C. Ho The Mass Function of Active Black Holes in the Local Universe, E. C. Moran, K. Shahinyan, H. R. Sugarman, D. O. Vélez and M. Eracleous Black Holes At the Centers of Nearby Dwarf Galaxies}. Table 2 of ~\citet{mercuza} displays a list of IMBH candidates found with these relations.

Another method to estimate large black hole masses is reverberation mapping, which estimates black hole virial masses by studying their nearby gasses' velocities \cite{reverberation mapping}. Reverberation mapping has shown to estimate both supermassive black hole and IMBH virial masses successfully. However, this method fails to measure the systematic uncertainties on the low-mass IMBH masses ~\cite{}. Additionally, cloud-variability and the possibility of anisotropic emission deride the credibility of IMBH candidates found with reverberation mapping.

The most promising IMBH candidates found with electromagnetic-spectra are Ultra-Luminous X-ray sources (with luminosities $> 10^{39} \text{erg s}^{-1}$). While stellar mass black holes are too small to generate such luminosities with Eddington accretion, Intermediate mass black holes can accrete at rates large enough to account for the high luminosities \cite{}. The drawback to this approach is the assumption that super-Eddington accretion is impossible and that X-ray emission from the sources is isotropic.

Finally, another channel to measure BH masses is from gravitational waves. Identifying GW from mergers of compact binaries. Ground-based gravitational-wave limits on the low end of this mass regime are continuing to improve. LIGO found 190521, can find more if we are more sensitive, and can better harness data.

IMBH hard to find as very similar to short-duration instrumental noise transients known as glitches~\cite{ pycbc_short_duration_transients, pe_with_glitch, blip_glitches}. These glitches can mimic astrophysical signals and hence decrease the significance of true gravitational-wave events.

One method to account for glitches while searching for gravitational-waves from coalescing compact binaries (CBCs) is by utilising the astrophysical Bayesian odds~\cite{bci, BCR1, BCR2, bcr_gw151216, bayesian_odds}. A true Bayesian odds, calculated without using bootstrap techniques, can provide more accurate estimate of significance that does note depend on the search pipeline~\cite{BCR2, bcr_gw151216,  bayesian_odds}.
In this paper, we utilise a Bayesian method, called the Bayesian Coherence Ratio \bcr~\cite{BCR1}, to rank the candidate gravitational-wave signals from high-mass compact binary coalescences (systems with total masses in the range of $55-500\ \msun$) in the detector data recorded during O2. Although the \bcr, utilising bootstrap techniques, does not provide the true Bayesian odds,  it utilizes Bayesian evidences which describe the explicit probability of data under the hypothesis that it contains coherent signals vs incoherent glitches.

@@

The \bcr is a Bayesian odds-ratio like the above, of a coherent signal hypotheses $\mathcal{H}_S$ and an incoherent instrumental feature hypothesis $\mathcal{H}_I$ (the null-hypotheses) for a network of $D$ detectors. $\mathcal{H_I}$ states that each detector $i$ has either pure stationary Gaussian noise $\mathcal{H}_N$ or Gaussian noise and an incoherent noise transient (glitch). Taking $\mathcal{H}_G$, $Z^S$, $Z^G_i$ and $Z^N_i$ as the Bayesian evidences (marginalised likelihoods) for $\mathcal{H}_S$, $\mathcal{H}_N$, and $\mathcal{H}_G$, the \bcr is given by
\begin{equation}
\label{eq:bcr}
\bcr = \frac{\alpha Z^S}{\prod\limits^D_{i=1} \ [\beta Z^G_i + (1-\beta)Z^N_i]}\ ,
\end{equation}
where $\alpha$ and $\beta$ are the prior-odds of obtaining a signal or a glitch from a stretch of data. The prior-odds can be defined more explicitly as 
\begin{itemize}
    \item $\alpha=P(\mathcal{H}_S)/P(\mathcal{H}_I)$, the prior-odds for obtaining a coherent signal versus an incoherent instrumental feature.
    \item $\beta=P(\mathcal{H}_G| \mathcal{H}_I)$, the prior-odds for obtaining a glitch assuming there is an incoherent instrumental feature.
\end{itemize}

@@

\subsection{New Candidate Events}
Although no clear detections are made with the \bcr, a marginal-candidate 170222 has been discovered with a $\pastrobcr\sim0.5$. This candidate has its similar masses when compared to those of GWTC-1. The remaining coherent trigger candidates all have $\pastrobcr\ll0.5$ making them unlikely to originate from astrophysical sources. \avi{wjat more can I say about this?}

@@

>>>>

%%%%%%---SECTIONS---%%%%%%%%%%%%
\section{Introduction}
In the 1970s, X-ray and optical observations from the Cyg-X1 binary star system led to the discovery of the first stellar-mass black hole~\cite{Webster:1972:Natur}. Two years later, in 1974, radio observation of our galaxy center demonstrated the presence of a supermassive black hole -- Sgr A$^*$~\cite{Balick:1974:ApJ}. In the half-century since then, astronomers accumulated more definitive evidence that have established the  existence of stellar-mass black holes $\text{M}_\text{BH} < 10^{2} \ \msun$ and supermassive black holes, $\text{M}_\text{BH} > 10^{5} \ \msun$ in our universe (e.g. ~\cite{Ghez:1998:ApJ, Genzel:2010:RvMP, Abbott:2019:PhRvX, EventHorizonTelescopeCollaboration:2019:ApJL, Abbott:2020:arXiv}). However, a gap in between the mass-ranges of these two categories has formed: there is a deficiency of observational evidence for black holes in the intermediate-mass range $10^{2} - 10^{5}\ \text{M}_{\odot}$. These intermediate-mass black holes must have existed to form the supermassive black holes detected to exist in the early universe~\cite{Banados:2018:Natur}. However, we only have a handful of promising candidates~\cite{Greene:2004:ApJ, Graham:2013:ApJ, Mezcua:2017:IJMPD, Koliopanos:2017:mbhe, Lin:2020:ApJL, Greene:2020:ARA&A}, and one confirmed intermediate-mass black hole, the remnant of the gravitational wave event GW190521 with a mass of $142^{+28}_{-16}\ \msun$~\cite{Abbott:2020:PhRvL}. The concrete discovery of more intermediate-mass black holes will bridge the observational gap and illuminate our understanding of galaxy and supermassive black hole formation.

The two main proposed formation channels for Intermediate-mass black holes are the light-seed and the heavy-seed channels. The light-seed channel involves dynamic interactions of low-mass seeds, e.g. hierarchical mergers of stellar-mass black holes~\cite{}, runaway collisions of stars~\cite{}.  The heavy-seed channel involves the direct collapse of high-mass seeds, e.g. the direct collapse of giant metal-free Population III stars~\cite{}, early protogalaxies' collapse~\cite{}.  Although promising pathways, there are challenging obstacles both scenarios must overcome to pathways that can form intermediate-mass black holes.

Hierarchical mergers in active galactic nuclei migration-traps, globular and stellar clusters may be a feasible light-seed method for black holes to accumulate mass and reach the intermediate-mass range. \rory{\sout{However, when black holes merge, they can receive recoil-kicks up to $\sim$2,500 km/s~\cite{}. Hence, the merger remnants require their recoil kick to be less than its environment's escape velocity to undergo successive mergers. This tight constraint makes hierarchical mergers a challenging scenario.}

One proposed heave-seed pathway involves the direct collapse of massive, metal-free Population III stars. Typically, massive stars $> 100 \ \msun$ can eject more than half of their mass in the form of radiatively driven stellar winds~\cite{}. However, at zero metallicity, mass loss through stellar winds is negligible \cite{Kudritzki 2000}. Consequently,  Population III stars have the opportunity to complete their life-cycles without losing a significant portion of their mass. Moreover, Population III stars with masses above $230 \ \msun$ will avoid electron-positron pair-instability. Therefore, oxygen and silicon burning will not drive an explosion, allowing the star to undergo a complete collapse to leave behind an intermediate-mass black hole remnant with a mass of $120-280 \ \msun$, dependant on the initial star mass \cite{Bond et al. 1984}. As the fraction of Population III stars with masses above $230 \ \msun$ is uncertain, the number of expected intermediate-mass black holes from these stars is also ambiguous~\cite{Coleman Miller, Formation Scenarios for Intermediate-Mass Black Holes}.

To verify and determine details on which formation channel is favoured, a catalog of observed intermediate-black holes and their masses is required. However, measuring the mass of black holes can be challenging.

\rory{\sout{It is possible to measure the mass of nearby supermassive black holes by directly observing the dynamic motion of stars nearby the supermassive black hole (such as Sgr A$^*$~). However, this dynamic measurement method is challenging to utilise for  intermediate-mass black holes as their sphere of influence is too small to resolve motions~\cite{}. Additionally, dynamic-measurements that have alluded to  intermediate-mass candidates, e.g. the candidate at the centre of the globular cluster 47 Tucanae~\cite{kiziltan 2017}, can also be attributed to simpler models without a central  intermediate-mass black hole~\cite{freire 2017}. }}

\rory{\sout{In a similar vein, although the M-$\sigma$ and M-L relations (relations between the spheroidal bulge of a galaxy's velocity dispersion $\sigma$,  luminosity L, and its central black hole's mass M) can estimate the mass of supermassive black holes~\cite{Ferrarese & Merritt 2000 and Gebhardt et al. 2000}, they cannot accurately estimate the intermediate-masses. The M-$\sigma$ and M-L relations are not calibrated for low-mass galaxies that might host an central intermediate-mass black hole~\cite{J. E. Greene and L. C. Ho The Mass Function of Active Black Holes in the Local Universe, E. C. Moran, K. Shahinyan, H. R. Sugarman, D. O. Vélez and M. Eracleous Black Holes At the Centers of Nearby Dwarf Galaxies}. Table 2 of~\citet{mercuza} displays a list of intermediate-mass candidates found with these relations. }}

\rory{\sout{Another method to estimate large black hole masses is with reverberation mapping, which estimates black hole virial masses by studying their nearby gasses' velocities~\cite{reverberation mapping}. Reverberation mapping has shown to estimate both supermassive black hole and intermediate-mass black hole virial masses successfully. However, this method fails to measure the systematic uncertainties on the lower-end of the intermediate-mass spectrum~\cite{}. Additionally, cloud-variability and the possibility of anisotropic emission deride the credibility of intermediate-mass candidates found with reverberation mapping~\cite{}. }}

The most promising intermediate-mass candidates found with electromagnetic-spectra are Ultra-Luminous X-ray sources with luminosities $> 10^{39} \text{erg s}^{-1}$ (e.g. HLX-X1)~\cite{}. \rory{\sout{While stellar mass black holes are too small to generate such luminosities with Eddington accretion, intermediate mass black holes can accrete at rates large enough to account for the high luminosities~\cite{}. However, some argue that some stellar mass black holes may exceed the Eddington limit~\cite{}, and others argue that the X-ray emission might be from a collimated jet leading to an inaccurate mass calculation~\cite{}.}}

\Finally, \rory{\sout{another channel to measure black hole masses is by identifying}} Gravitational waves from compact binaries coalesces (CBCs) \rory{are clean channels to measure BH masses} \rory{\sout{ inferring their source parameters~\cite{}.}} Mergers involving either a binary system with at least one intermediate-mass component, or a system whose merger results in an intermediate-mass remnant will have large initial total masses $\text{M}_{T}$. As a binary's $\text{M}_{T}$ is associated with its merger frequency, $f\sim \text{M}_T^-1$, systems with large $\text{M}_{T}$ have very low merger frequencies $f < 100\ \text{Hz}$. Hence, ground based gravitational wave detectors ($\sim 10 - 1000\ \text{Hz}$) are sensitive to the last milliseconds of merging systems with $100 \msun < \text{M}_{T} < 500 \msun$, while space-based detectors ($\sim 0.1 - 10\ \text{Hz}$) can study the full signals of merging systems with $10^4 \msun < \text{M}_{T} < 10^7 \msun$.

Because of high-total mass systems' short-durations in ground-based gravitational wave data steams, handling data quality is critical to their detection. The low-frequency ranges of ground-based observatories are plagued with incoherent non-stationary terrestrial artefacts, called glitches~\cite{ pycbc_short_duration_transients, pe_with_glitch, blip_glitches}. Some glitches, similar to signals from high total mass mergers, last for a fraction of a second, making them difficult to distinguish from the signals. These glitches that mimic astrophysical signals can severely decrease the confidence in detection of true gravitational-waves from high total mass mergers.

Although a significant fraction of the glitches can be removed by testing them for coherence amongst various ground based detectors and performing matched-filtering, these methods are insufficient to remove all the glitches. One method to account for more glitches while searching for high total mass CBC gravitational-waves is by utilising the astrophysical Bayesian odds~\cite{bci, BCR1, BCR2, bcr_gw151216, bayesian_odds}. A true Bayesian odds, calculated without using bootstrap techniques, can provide more accurate estimate of significance that does note depend on the search pipeline~\cite{BCR2, bcr_gw151216,  bayesian_odds}. In this paper, we utilise a Bayesian method, called the Bayesian Coherence Ratio \bcr~\cite{BCR1}, to rank the candidate gravitational-wave signals from high-mass compact binary coalescences (systems with total masses in the range of $55-500\ \msun$) in the detector data recorded during O2. Although the \bcr, utilising bootstrap techniques, does not provide the true Bayesian odds,  it utilizes Bayesian evidences which describe the explicit probability of data under the hypothesis that it contains coherent signals vs incoherent glitches.

\avi{Need to discuss IAS, PyCBC etc here as well}

@@

The \bcr is a Bayesian odds-ratio like the above, of a coherent signal hypotheses $\mathcal{H}_S$ and an incoherent instrumental feature hypothesis $\mathcal{H}_I$ (the null-hypotheses) for a network of $D$ detectors. $\mathcal{H_I}$ states that each detector $i$ has either pure stationary Gaussian noise $\mathcal{H}_N$ or Gaussian noise and an incoherent noise transient (glitch). Taking $Z^S$, $Z^G_i$ and $Z^N_i$ as the Bayesian evidences (marginalised likelihoods, definitions in Appendix~\ref{apdx} for $\mathcal{H}_S$, $\mathcal{H}_N$, and $\mathcal{H}_G$, the \bcr is given by
\begin{equation}
\label{eq:bcr}
\bcr = \frac{\alpha Z^S}{\prod\limits^D_{i=1} \ [\beta Z^G_i + (1-\beta)Z^N_i]}\ ,
\end{equation}
where $\alpha$ and $\beta$ are the prior-odds of obtaining a signal or a glitch from a stretch of data. The prior-odds can be defined more explicitly as 
\begin{itemize}
    \item $\alpha=P(\mathcal{H}_S)/P(\mathcal{H}_I)$, the prior-odds for obtaining a coherent signal versus an incoherent instrumental feature.
    \item $\beta=P(\mathcal{H}_G| \mathcal{H}_I)$, the prior-odds for obtaining a glitch assuming there is an incoherent instrumental feature.
\end{itemize}

@@

\subsection{New Candidate Events}
Although no clear detections are made with the \bcr, a marginal-candidate 170222 has been discovered with a $\pastrobcr\sim0.5$. This candidate has its similar masses when compared to those of GWTC-1. The remaining coherent trigger candidates all have $\pastrobcr\ll0.5$ making them unlikely to originate from astrophysical sources.

\avi{mass and spin estimates, is it a vanilla bbh}

@@

% Gravitational waves from intermediate-mass black holes will Current ground-based gravitational-wave observatories like LIGO can probe the lower-end of the intermediate-mass range, u, while future space based detectors will be able to study the mergers of large limits on the low end of this mass regime are continuing to improve. LIGO found 190521, can find more if we are more sensitive, and can better harness data.

% Most of what we know about intermediate mass black holes so far has been informed from electromagnetic observations. Gravitational waves provide a complementary lens to study these massive objects. In the future, LISA will fill in this gap, but we can begin now with LIGO... In this paper, we
